---
title: "Leaflet maps"
author: "David Wilkinson"
date: "10 January 2018"
output: 
  html_document:
    css: lodestar.css
    toc: yes
    toc_float:
      collapsed: no
      toc_depth: 4
---

<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #208000;
}
</style>

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(leaflet)
library(maps)
library(viridis)
library(RColorBrewer)
library(mapr)
library(zoon)

data(quakes)
```

<hr class="double">

## Leaflet Maps

Leaflet maps are interactive maps made using the `Leaflet.js` JavaScript library with the help of the `leaflet` R package. These maps are really effective for "end-of-project" uses like presentations or in digital publications like blogs and websites. This is their most common usage, but they also provide some unique approaches to "start-of-project" tasks like data exploration.

An [excellent guide](https://rstudio.github.io/leaflet/) for Leaflet maps already exists, so instead of re-inventing the wheel this guide will step through the basics and then cover some ecology-related uses

<hr class="small">

## The Basics

Leaflet maps are built with a syntax similar to the *grammar of graphics* approach of `ggplot2`. You start with an empty plot, add a base-layer on top, maybe add some more layers on top, possibly add some markers, do any customising you want, and run the code. 

One obvious difference is that `leaflet` uses the `%>%` pipe from `magrittr` instead of the `+` in `ggplot2`, but the usage is the same.

A basic Leaflet map looks like this:

```{r}
leaflet()
```


This is equivalent to an empty `ggplot()` graph.

<hr class="small">

### Base Maps

The first thing you will normally want to do is add a basemap with the `addTiles()` function. By default this is the `OpenStreetMap`. While we're at it, lets also use the `setView()` function so we're not looking at the whole globe by default.

```{r}
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 144.9612, lat = -37.7964, zoom = 13)
```


There are also a lot of freely-available third-party layers that can be accessed with the `addProviderTiles()` function. Examples of all of these can be found [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html). Some examples:

```{r}
leaflet() %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  setView(lng = 15.2663, lat = 4.4419, zoom = 5)
```

```{r}
leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  setView(lng = -77.7812, lat = 21.5218, zoom = 4)
```

```{r}
leaflet() %>%
  addProviderTiles(providers$Stamen.Toner) %>%
  setView(lng = -76.6122, lat = 39.2904, zoom = 10)
```


There are also third-party maps that cost nothing, but require registration to get an api key. Adding these goes back to the `addTiles()` function. In fact, anything you want to add that isn't one of the available provider tiles is done with `addTiles()`.

```{r}
leaflet() %>%
  addTiles("http://{s}.tile.thunderforest.com/spinal-map/{z}/{x}/{y}.png?apikey=276ae9b671ed49e4a663b0edd985b2d1") %>%
  setView(lng = 4.30535, lat = 50.8549541, zoom = 5)
```


Once you have a basemap picked out, you're going to want to add stuff on top.

### Polygons

One of the most common uses for leaflet maps that you will see online is chloropleth maps. These maps shade areas based on a measurement of some statistical value like per-capita income or housing prices. These are done by adding polygons to a map using the `addPolygons()` function. Here, lets colour the states of the contiguous USA with a random draw (because I don't have anything better to hand that's simple):

```{r}
mapStates <- map("state", fill = TRUE, plot = FALSE)

leaflet(data = mapStates) %>% 
  addTiles() %>%
  addPolygons(fillColor = rainbow(50, alpha = NULL),
              stroke = FALSE)
```

### Markers

Another thing that can be added to Leaflet maps are point markers using `addMarkers()`. These can be as simple as icons on a map:

```{r}
# Show first 20 rows from the `quakes` dataset

leaflet(data = quakes[1:20,]) %>% 
  addTiles() %>%
  addMarkers(~long, ~lat)
```


To having pop-up information:

```{r}
leaflet(data = quakes[1:20,]) %>% 
  addTiles() %>%
  addMarkers(~long, ~lat, 
             popup = ~as.character(mag),
             label = ~as.character(mag))
```


Or custom icons with conditional colouring:

```{r}
df.20 <- quakes[1:20,]

getColor <- function(quakes) {
  sapply(quakes$mag, function(mag) {
  if(mag <= 4) {
    "green"
  } else if(mag <= 5) {
    "orange"
  } else {
    "red"
  } })
}

icons <- awesomeIcons(
  icon = 'fa-flag-checkered',
  iconColor = 'black',
  library = 'fa',
  markerColor = getColor(df.20)
)

leaflet(df.20) %>% addTiles() %>%
  addAwesomeMarkers(~long, ~lat, icon=icons, label=~as.character(mag))
```


If you have lots of points you can even cluster them:

```{r message=FALSE, warning=FALSE}
leaflet(quakes) %>%
  addTiles() %>%
  addMarkers(clusterOptions = markerClusterOptions())
```

<hr class="small">

## Some Ecological Examples

Lets set up some data exploration for a species distribution model. Here we add our covariate rasters using `addTile()`, add our presence points with `addCircleMarkers()`, and make use of `groups` to control what can be turned on/off simultaneously, can only be shown one at a time, and what is visible at the beginning.

```{r echo=FALSE, message=FALSE, warning=FALSE}
LoadModule("CarolinaWrenPO")
LoadModule("Bioclim")

data <- CarolinaWrenPO()

ras <- Bioclim(extent = c(-138.71, -52.58, 18.15, 54.95),
               layers = 1:10)
```

```{r message=FALSE, warning=FALSE}
## Set up map + base layers

map <- leaflet(data) %>%
    addTiles(group = "OpenStreetMap") %>%
    addProviderTiles(provider = 'Esri.WorldImagery',
                     group = 'Esri.WorldImagery')

## Set up group layers
  
overlay_groups <- c() # layers to "print"
hide_groups <- c() # layers to not have ticked by default
  
## Point colour palette
  
fill_pal <- colorFactor(c("#FFFFFF", "#808080", "#080808"),
                        domain = c('presence', 'background', 'absence'),
                        ordered = TRUE)
  
border_pal <- colorFactor(c("#F0F0F0", "#808080", "#080808"),
                          domain = c('presence', 'background', 'absence'),
                          ordered = TRUE)  
  
## Data points
  
group_name <- "Presence Points"  # give layer name

overlay_groups <- c(overlay_groups, group_name)  # add to overlay group
      
map <- addCircleMarkers(map = map,  # add circles to map
                        lng = data$longitude,  # longitude values
                        lat = data$latitude,  # latitude values
                        color = grey(0.4),  # border colour
                        fillColor = "#FFFFFF",  # fill colour
                        weight = 1,
                        opacity = 1,
                        fillOpacity = 1,
                        radius = 3.5,
                        group = group_name,
                        popup = paste('<b>',
                                      paste(toupper(substr("presence", 1, 1)),
                                            substr("presence", 2, nchar("presence")), sep=""),
                                      '</b>',
                                      '<br>Longitude:', data$longitude,
                                      '<br>Latitude:', data$latitude))

## Add points legend
  
map <- addLegend(map = map,
                 pal = fill_pal,
                 opacity = 0.8,
                 values = factor(c('presence', 'absence', 'background'),
                                 levels = c('presence', 'absence', 'background'),
                                 ordered = TRUE),
                 title = 'Data points')
  
## Add Covariate Rasters
  
for(i in seq(nlayers(ras))){
  
  ### layer names
  
  layer_name_tmp <- sprintf('names(ras)[%s]', i)  # get raster layer name
  layer_name <- eval(parse(text = layer_name_tmp))
  overlay_groups <- c(overlay_groups, layer_name)  # add layer to overlay group
    
  ### get raster colour palette
    
  ras_pal <- colorNumeric(viridis(10),    # set colour palette by viridis
                          domain = seq(minValue(ras[[i]]),
                                       maxValue(ras[[i]]),
                                       length.out = 10),
                          na.color = 'transparent')
  
  ### reproject ras[[i]], suppressing warnings
    
  suppressWarnings(ext <- raster::projectExtent(ras[[i]], crs = sp::CRS('+init=epsg:3857')))
  suppressWarnings(tmp_ras <- raster::projectRaster(ras[[i]], ext))
    
  ### add layer
    
  map <- addRasterImage(map = map,  # add layer to map
                        x = tmp_ras,
                        colors = ras_pal,
                        project = FALSE,
                        opacity = 0.8,
                        group = layer_name)
    
  ### add all bar first raster layer to hide group so default display is one raster
    
  if(i != 1){
    hide_groups <- c(hide_groups, layer_name)
    }
  }
  
## Add toggle for the layers
  
map <- addLayersControl(map = map,   # add toggle switch for layers
                        position = "topleft",
                        baseGroups = c('OpenStreetMap',
                                       'Esri.WorldImagery'),
                        overlayGroups = overlay_groups,
                        options = layersControlOptions(collapsed = FALSE))
  
map <- hideGroup(map = map,  # untick all raster layers except first
                 hide_groups)
  
map
```

**Add some more examples. Maybe something with reserve borders and polygons? The Measurement add-on for field site spacing? Transect distances?**